---
title: "Seurat clustering"
author: "Amy Gill"
date: "4/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data


```{r}
library(SingleCellExperiment)
library(Seurat)
library(tidyverse)

colorblind_palette <- c("#000000", "#D55E00",  "#56B4E9", "#E69F00", "#009E73", "#F0E442", "#0072B2",  "#CC79A7")

load("sce.rda")
```

## Create Seurat object from SingleCellExperiment

```{r}
tnbc_seurat <- as.Seurat(sce, counts = "normcounts", data = "normcounts")
```

```{r}
tnbc_seurat
```

## Regress out patient effect

In the original paper, clustering was performed after regressing out patient-specific effects. Therefore, I will do the same.

```{r}
tnbc_seurat <- ScaleData(tnbc_seurat, vars.to.regress = "patient_id", features = rownames(tnbc_seurat))
```


## Find variable features

```{r}
tnbc_seurat <- FindVariableFeatures(tnbc_seurat, selection.method = "vst", nfeatures = 2000)

top8 <- head(VariableFeatures(tnbc_seurat), 8)

p1 <- VariableFeaturePlot(tnbc_seurat)
p2 <- LabelPoints(plot = p1, points = top8, repel = TRUE, xnudge = 0.05, ynudge = 0.01)
p2
```

## PCA dimensional reduction

```{r}
tnbc_seurat <- RunPCA(tnbc_seurat, features = VariableFeatures(tnbc_seurat))
```
```{r}
ElbowPlot(tnbc_seurat, ndims = 50)
```

## Run tSNE and UMAP

```{r}
tnbc_seurat <- RunTSNE(tnbc_seurat, dims = 1:50)
tnbc_seurat <- RunUMAP(tnbc_seurat, dims = 1:50)
```

## Color tSNE and UMAP by patient

```{r}
DimPlot(tnbc_seurat, reduction = "tsne",
		group.by = "patient_id") +
	ggtitle("tSNE plot of patients after removing patient effect") +
	scale_color_manual(values = colorblind_palette[c(1:5,8)])
```


```{r}
DimPlot(tnbc_seurat, reduction = "umap",
		group.by = "patient_id") +
	ggtitle("UMAP plot of patients after removing patient effect") +
	scale_color_manual(values = colorblind_palette[c(1:5,8)])
```

## Find clusters

### Find neighbors

```{r}
tnbc_seurat <- FindNeighbors(tnbc_seurat, dims = 1:50)
```

### Testing clustering resolutions

```{r}
# run clustering with a range of resolution parameters
rez <- seq(0.4, 1.2, 0.1)
clust <- sapply(rez, function(x){
	trez <- FindClusters(tnbc_seurat, resolution = x)
	length(table(Idents(trez)))
})
rez_clust <- data.frame(rez, clust)
# scatter plot and curve fit for number of clusters versus Louvain resoluton parameter
rez_clust %>%
	ggplot(aes(rez, clust)) +
	geom_point() + 
	geom_line() +
	xlab("Resolution parameter of Louvain clustering") +
	ylab("Number of clusters") +
	ggtitle("Clusters identified at different resolutions by the Louvain algorithm") +
	ylim(c(5,20))
```
At resolutions greater than or equal to 1, there are 18 clusters. Moving forward, I use a resolution of 1.

```{r}
tnbc_seurat <- FindClusters(tnbc_seurat, resolution = 1)
```

```{r}
table(tnbc_seurat$seurat_clusters)
```

## tSNE


```{r}
DimPlot(tnbc_seurat, reduction = "tsne",
		group.by = "seurat_clusters",
		label = TRUE, label.size = 5, repel = TRUE) +
	ggtitle("tSNE plot of Louvain clusters")
```


## UMAP


```{r}
DimPlot(tnbc_seurat, reduction = "umap",
		group.by = "seurat_clusters",
		label = TRUE, label.size = 5, repel = TRUE) +
	ggtitle("UMAP plot of Louvain clusters")
```


## Save updated Seurat and SingleCellExperiment objects

```{r}
# save updated tnbc_seurat object
save(tnbc_seurat, file = "tnbc_seurat.rda")

# create updated sce object
sce_post_seurat <- as.SingleCellExperiment(tnbc_seurat)
save(sce_post_seurat, file = "sce_post_seurat.rda")
```


## Differentially expressed genes in Seurat clusters

```{r}
seurat_markers <- FindAllMarkers(tnbc_seurat, logfc.threshold = 0.5, slot = "scale.data")

# keep only differentially expressed markers with an adjusted pval less than .05
seurat_markers <- seurat_markers %>%
	filter(p_val_adj <= .05)

save(seurat_markers, file = "seurat_markers.rda")
```

```{r}
# number of differentially expressed genes for each cluster
seurat_markers %>%
	group_by(cluster) %>%
	summarize(n = n())
```


```{r}
top5 <- seurat_markers %>%
	group_by(cluster) %>%
	top_n(5, wt = avg_log2FC)

DoHeatmap(tnbc_seurat, features = top5$gene) + NoLegend()
```

```{r}
top10 <- seurat_markers %>%
	group_by(cluster) %>%
	top_n(10, wt = avg_log2FC)

DoHeatmap(tnbc_seurat, features = top10$gene) + NoLegend() +
	theme(axis.text.y = element_blank())
```






