---
title: "Literature-based cell typing"
author: "Amy Gill"
date: "4/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load files and libraries

```{r}
library(tidyverse)
library(stringr)
library(SingleCellExperiment)

colorblind_palette <- c("#000000", "#D55E00",  "#56B4E9", "#E69F00", "#009E73", "#F0E442", "#0072B2",  "#CC79A7")

load("cell_type_markers.rda")   # output from import-marker-gene-list.Rmd
load("sce_post_seurat.rda")    # output from seurat-clustering.Rmd
```

## Expression of all expert-based marker genes above cutoff threshold

All rules were derived from the supplementary material, p35-36.

From the supplement, "Expression" = at or above a value of 1

### Pull normalized counts into separate matrix

Goal: shorthand for easy calling of the count matrix

```{r}
ct <- logcounts(sce_post_seurat)
```


### Pull colData into separate editable data frame

Goal: use to track expr of various markers in the cells without modifying colData in place

```{r}
df <- colData(sce_post_seurat)
df
```

### Check expression of all marker genes above a threshold of 1

```{r}
for (i in 1:nrow(cell_type_markers)) {
	gene <- cell_type_markers$marker[i]
	df[, gene] <- ct[gene,] > 1
}

df
```



## Find epithelial cells

### Pull epithelial markers and strong epithelial markers

```{r}
epi_markers <- cell_type_markers %>%
	filter(cell_type == "epithelial") %>%
	pull(marker)

epi_markers
```

```{r}
strong_epi_markers <- c("EPCAM", "KRT8", "KRT18", "KRT19")
strong_epi_markers
```

### Case 1: Cell expresses at least 2 epithelial markers

```{r}
# calculate number of epithelial genes expressed above a threshold of 1
epi <- df[, epi_markers]
epi_counts <- rowSums(as.matrix(epi))

hist(epi_counts)
```
```{r}
# mark cells as epithelial if they pass case 1
df$epithelial <- epi_counts >= 2
table(df$epithelial)
```

### Case 2: Cells highly express only one strong epithelial marker

Cells express only one of the strongest markers (EPCAM, KRT8, KRT18, KRT19) with expression higher than 50% of the cells for that patient, for that marker.

```{r}
# inspect strong_epi_markers in cells with an epi_count of 1
strong_epi <- df[epi_counts == 1, strong_epi_markers]

# remove cells without significant expr of the strong_epi markers from strong_epi
strong_epi <- strong_epi[rowSums(as.matrix(strong_epi)) > 0,]


# from strong_epi, extract the single expressed marker whose expression must be compared to median for that pt
marker_test <- data.frame(
	sample_id = rownames(strong_epi),
	patient_id = df[df$sample_id %in% rownames(strong_epi), "patient_id"],
	marker = case_when(
		strong_epi$EPCAM ~ "EPCAM",
		strong_epi$KRT8 ~ "KRT8",
		strong_epi$KRT18 ~ "KRT18",
		strong_epi$KRT19 ~ "KRT19"
	)
)
marker_test
```

```{r}
# create data frame to store patient-specific median expression for strong_epi_markers
pt_epi_df <- data.frame(patient_id = names(table(sce_post_seurat$patient_id)))

for (i in 1:length(strong_epi_markers)) {
	# set column name to Marker_50
	colname <- paste0(strong_epi_markers[i], "_50")
	
	# add median expression for that marker for each patient to pt_epi_df
	pt_epi_df[,colname] <- sapply(1:nrow(pt_epi_df), function(x){
		pt <- pt_epi_df$patient_id[x]
		pt_expr <- sce_post_seurat[, sce_post_seurat$patient_id == pt]
		# find median expression for that patient for that marker
		median(logcounts(pt_expr[strong_epi_markers[i],]))
	})
}

pt_epi_df
```


```{r}
marker_test$expr_threshold <- sapply(1:nrow(marker_test), function(i){
	marker <- paste0(marker_test$marker[i],"_50")
	temp <- pt_epi_df %>%
		filter(patient_id == marker_test$patient_id[i])
	
	temp[, marker]
})

marker_test
```

```{r}
marker_test$marker_expr <- sapply(1:nrow(marker_test), function(i){
	temp <- marker_test[i,]
	gene <- temp$marker
	sample_id <- temp$sample_id
	logcounts(sce_post_seurat[gene, sample_id])
})

marker_test
```


```{r}
# find sample IDs for cells that pass epithelial case 2
epi2_samples <- marker_test %>%
	filter(marker_expr > expr_threshold) %>%
	pull(sample_id)

# mark cells as epithelial if they pass case 2 OR case 1
df$epithelial <- ifelse(rownames(df) %in% epi2_samples, TRUE, df$epithelial)
table(df$epithelial)
```


## Pull non-epithelial markers

### Pull immune markers 

```{r}
# generic immune marker PTPRC
ptprc <- "PTPRC"

# T cell markers
tcell_markers <- cell_type_markers %>%
	filter(cell_subtype == "T cell") %>%
	pull(marker)

# B cell markers
bcell_markers <- cell_type_markers %>%
	filter(cell_subtype == "B cell") %>%
	pull(marker)

# macrophage markers
macro_markers <- cell_type_markers %>%
	filter(cell_subtype == "macrophage") %>%
	pull(marker)
```

### T cell gene expression count

```{r}
# calculate number of epithelial genes expressed above a threshold of 1
tcell <- df[, tcell_markers]
tcell_counts <- rowSums(as.matrix(tcell))

hist(tcell_counts)
```

### B cell gene expression count

```{r}
# calculate number of T cell genes expressed above a threshold of 1
bcell <- df[, bcell_markers]
bcell_counts <- rowSums(as.matrix(bcell))

hist(bcell_counts)
```

### Macrophage gene expression count

```{r}
macro <- df[, macro_markers]
macro_counts <- rowSums(as.matrix(macro))

hist(macro_counts)
```

```{r}
stroma_markers <- cell_type_markers %>%
	filter(cell_type == "stroma") %>%
	pull(marker)

endo_markers <- cell_type_markers %>%
	filter(cell_type == "endothelial") %>%
	pull(marker)
```

### Stromal gene expression count

```{r}
stroma <- df[, stroma_markers]
stroma_counts <- rowSums(as.matrix(stroma))

hist(stroma_counts)
```

### Endothelial gene expression count

```{r}
endo <- df[, endo_markers]
endo_counts <- rowSums(as.matrix(endo))

hist(endo_counts)
```

```{r}
non_epi_df <- data.frame(sample_id = df$sample_id,
						 ptprc = df[,ptprc],
						 tcell_counts = tcell_counts,
						 bcell_counts = bcell_counts,
						 macro_counts = macro_counts,
						 stroma_counts = stroma_counts,
						 endo_counts = endo_counts)

non_epi_df
```

## Classify immune cells

### T cells

#### T cell case 1: only T cell immune markers

```{r}
# at least 2 T cell markers and no other immune markers
tcell_id_1 <- non_epi_df %>%
	filter(tcell_counts >= 2,
#		   epi_counts == 0,
		   bcell_counts == 0,
		   macro_counts == 0,
#		   stroma_counts == 0,
#		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### T cell case 2: PTPRC and only T cell immune markers (at least 1)

```{r}
# PTPRC and 1 T cell marker with no other immune markers
tcell_id_2 <- non_epi_df %>%
	filter(ptprc == TRUE,
		   tcell_counts >= 1,
#		   epi_counts == 0,
		   bcell_counts == 0,
		   macro_counts == 0,
		   stroma_counts == 0,
		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### T cell case 3: At least 3 T cell markers and no more than 1 other immune markers

```{r}
# at least 3 T cell markers and no more than 1 other immune cell marker
tcell_id_3 <- non_epi_df %>%
	filter(tcell_counts >= 3,
		   bcell_counts + macro_counts <= 1,
#		   (bcell_counts == 1 & macro_counts == 0) | (bcell_counts == 0 & macro_counts == 1),
#		   epi_counts == 0,
#		   stroma_counts == 0,
#		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### Label all T cells

```{r}
tcell_id <- c(tcell_id_1, tcell_id_2, tcell_id_3)

df$tcell <- df$sample_id %in% tcell_id

table(df$tcell)
```

### B cells

#### B cell case 1: only B cell immune markers

```{r}
# at least 2 B cell markers and no other immune markers
bcell_id_1 <- non_epi_df %>%
	filter(bcell_counts >= 2,
		   tcell_counts == 0,
		   macro_counts == 0,
#		   epi_counts == 0,
#		   stroma_counts == 0,
#		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### B cell case 2: PTPRC and only B cell immune markers (at least 1)

```{r}
# PTPRC and 1 B cell marker with no other immune markers
bcell_id_2 <- non_epi_df %>%
	filter(ptprc == TRUE,
		   bcell_counts >= 1,
		   tcell_counts == 0,
		   macro_counts == 0,
#		   epi_counts == 0,
		   stroma_counts == 0,
		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### B cell case 3: At least 3 B cell markers and no more than 1 other immune markers

```{r}
# at least 3 B cell markers and no more than 1 other immune cell marker
bcell_id_3 <- non_epi_df %>%
	filter(bcell_counts >= 3,
		   tcell_counts + macro_counts <= 1,
#		   (tcell_counts == 1 & macro_counts == 0) | (tcell_counts == 0 & macro_counts == 1),
#		   epi_counts == 0,
#		   stroma_counts == 0,
#		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### Label all B cells

```{r}
bcell_id <- c(bcell_id_1, bcell_id_2, bcell_id_3)

df$bcell <- df$sample_id %in% bcell_id

table(df$bcell)
```

### Macrophages

#### Macrophage case 1: only macrophage immune markers

```{r}
# at least 2 macrophage markers and no other immune markers
macro_id_1 <- non_epi_df %>%
	filter(macro_counts >= 2,
		   tcell_counts == 0,
		   bcell_counts == 0,
#		   epi_counts == 0,
#		   stroma_counts == 0,
#		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### Macrophage case 2: PTPRC and only macrophage immune markers (at least 1)

```{r}
# PTPRC and 1 macrophage marker with no other immune markers
macro_id_2 <- non_epi_df %>%
	filter(ptprc == TRUE,
		   macro_counts >= 1,
		   tcell_counts == 0,
		   bcell_counts == 0,
#		   epi_counts == 0,
		   stroma_counts == 0,
		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### Macrophage case 3: At least 3 macrophage markers and no more than 1 other immune markers

```{r}
# at least 3 macrophage markers and no more than 1 other immune cell marker
macro_id_3 <- non_epi_df %>%
	filter(macro_counts >= 3,
		   tcell_counts + bcell_counts <= 1,
#		   epi_counts == 0,
#		   stroma_counts == 0,
#		   endo_counts == 0
) %>%
	pull(sample_id)
```

#### Label all macrophages

```{r}
macro_id <- c(macro_id_1, macro_id_2, macro_id_3)

df$macro <- df$sample_id %in% macro_id

table(df$macro)
```


## Classify stromal and endothelial cells

### Stromal cells

#### Stromal cell case 1: Only stromal markers

```{r}
stroma_id_1 <- non_epi_df %>%
	filter(stroma_counts > 0,
		   endo_counts == 0,
		   epi_counts == 0,
		   tcell_counts == 0,
		   bcell_counts == 0,
		   macro_counts == 0) %>%
	pull(sample_id)
```


#### Stromal cell case 2: At least 3 stromal markers and no more than 1 endothelial marker

```{r}
stroma_id_2 <- non_epi_df %>%
	filter(stroma_counts >= 3,
		   endo_counts <= 1) %>%
	pull(sample_id)
```

#### Label all stromal cells

```{r}
stroma_id <- unique(c(stroma_id_1, stroma_id_2))

df$stromal <- df$sample_id %in% stroma_id

table(df$stromal)
```

### Endothelial cells

#### Endothelial cell case 1: Only endothelial markers

```{r}
endo_id_1 <- non_epi_df %>%
	filter(endo_counts > 0,
		   epi_counts == 0,
		   stroma_counts == 0,
		   tcell_counts == 0,
		   bcell_counts == 0,
		   macro_counts == 0) %>%
	pull(sample_id)
```


#### Endothelial cell case 2: At least 3 endothelial markers and no more than 1 stromal marker

```{r}
endo_id_2 <- non_epi_df %>%
	filter(endo_counts >= 3,
		   stroma_counts <= 1) %>%
	pull(sample_id)
```

#### Label all endothelial cells

```{r}
endo_id <- unique(c(endo_id_1, endo_id_2))

df$endothelial <- df$sample_id %in% endo_id

table(df$endothelial)
```

## Decide cell type 

```{r}
type_df <- as.data.frame(df) %>%
	select(sample_id, patient_id, epithelial, tcell, bcell, macro, stromal, endothelial)


epi_only_id <- type_df %>%
	filter(epithelial, !tcell, !bcell, !macro, !endothelial) %>%  # don't remove things classified as epi + stroma (EMT)
	pull(sample_id)

t_only_id <- type_df %>%
	filter(tcell, !epithelial, !bcell, !macro, !stromal, !endothelial) %>%
	pull(sample_id)

b_only_id <- type_df %>%
	filter(bcell, !epithelial, !tcell, !macro, !stromal, !endothelial) %>%
	pull(sample_id)

macro_only_id <- type_df %>%
	filter(macro, !epithelial, !tcell, !bcell, !stromal, !endothelial) %>%
	pull(sample_id)

stromal_only_id <- type_df %>%
	filter(stromal, !epithelial, !tcell, !bcell, !macro, !endothelial) %>%
	pull(sample_id)

endo_only_id <- type_df %>%
	filter(endothelial, !epithelial, !tcell, !bcell, !macro, !stromal) %>%
	pull(sample_id)

unknown_id <- type_df %>%
	filter(!epithelial, !tcell, !bcell, !macro, !stromal, !endothelial) %>%
	pull(sample_id)

undecided_id <- type_df %>%
	filter(!(sample_id %in% c(epi_only_id, t_only_id, b_only_id, macro_only_id,
							  stromal_only_id, endo_only_id, unknown_id))) %>%
	pull(sample_id)
```

```{r}
lit_cell_type <- type_df %>%
	mutate(cell_type = case_when(
		sample_id %in% epi_only_id ~ "epithelial",
		sample_id %in% t_only_id ~ "T cell",
		sample_id %in% b_only_id ~ "B cell",
		sample_id %in% macro_only_id ~ "macrophage",
		sample_id %in% stromal_only_id ~ "stromal",
		sample_id %in% endo_only_id ~ "endothelial",
		sample_id %in% unknown_id ~ "unknown",
		sample_id %in% undecided_id ~ "undecided"
	)) %>%
	pull(cell_type)



table(lit_cell_type)
```

```{r}
sce_post_seurat$lit_cell_type <- lit_cell_type
```

```{r}
save(sce_post_seurat, file = "sce_post_seurat.rda")
```

## Cell type proportions bar graph

```{r}
as.data.frame(colData(sce_post_seurat)) %>%
	group_by(patient_id, lit_cell_type) %>%
	summarize(n=n()) %>%
	ggplot(aes(patient_id, n, fill = lit_cell_type)) +
	geom_bar(stat = "identity", position = "fill") +
	xlab("Patient ID") +
	ylab("Proportion of cells") +
	scale_fill_manual (values = colorblind_palette)
```


```{r}
as.data.frame(colData(sce_post_seurat)) %>%
	group_by(patient_id, lit_cell_type) %>%
	summarize(n=n()) %>%
	ggplot(aes(patient_id, n, fill = lit_cell_type)) +
	geom_bar(stat = "identity") +
	scale_fill_manual(values = colorblind_palette) +
	xlab("Patient ID") +
	ylab("Number of cells")
```

