---
title: "SC3 consensus clustering"
author: "Amy Gill"
date: "5/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and data

```{r message = FALSE}
library(SingleCellExperiment)
library(SC3)
library(tidyverse)
library(Seurat)

load("sce.rda")    # normalized counts

load("sce_tpm.rda")    # raw counts for calculating dropouts

load("tnbc_seurat.rda")   # for visualization
```

## Format SingleCellExperiment for SC3

```{r}
sce
```

```{r}
sce_tpm
```

```{r}
# subset raw count assay to the filtered genes and filtered cells
sce_sc3 <- sce_tpm[rownames(sce),colnames(sce)]

# add normalized counts (slot is required to be named logcounts for compatibility with SC3)
assay(sce_sc3, "logcounts") <- assay(sce)

sce_sc3
```

```{r}
# clean workspace
rm(sce)
rm(sce_tpm)
```


## Perform SC3 stepwise for extra customization

### Prepare parameters

```{r}
# instead of performing sc3 as a single wrapper call (sc3(sce_sc3, ks = ...)),
# perform it piece-wise so k can be estimated with sc3_estimate_k

sce_sc3 <- sc3_prepare(sce_sc3)
```
```{r}
str(metadata(sce_sc3)$sc3)
```

### Estimate optimal k

The optimal value of k was estimated using the `sc3_estimate_k` function from SC3. "SC3 utilizes the Tracy-Widom theory on random matrices to estimate k"

```{r}
set.seed(42)   # I don't think the estimation is stochastic, but adding this just in case
sce_sc3 <- sc3_estimate_k(sce_sc3)
```
```{r}
# show estimated optimal k
metadata(sce_sc3)$sc3$k_estimation
```

### Calculate distances between cells

```{r}
sce_sc3 <- sc3_calc_dists(sce_sc3)
```

```{r}
names(metadata(sce_sc3)$sc3$distances)
```

### Transform distance matrix with PCA and graph Laplacian

```{r}
sce_sc3 <- sc3_calc_transfs(sce_sc3)
```

```{r}
names(metadata(sce_sc3)$sc3$transformations)
```

### Perform k-means clustering

Use 3 values of k: 6 (the number of input cell types), 18 (the number of clusters from Louvain clustering), and 29 (the optimal value calculated)

```{r}
set.seed(42)    # again, not sure this next step is stochastic, but it very well may be because k-means usually is
sce_sc3 <- sc3_kmeans(sce_sc3, ks = c(6, 18, 29))
```

```{r}
names(metadata(sce_sc3)$sc3$kmeans)
```

```{r}
# save object for now to retain k-means info, because that is overwritten in the consensus matrix calculation below
sce_kmeans <- sce_sc3
save(sce_kmeans, file = "sce_kmeans.rda")
```


### Calculate consensus matrix

```{r}
# check that there are no SC3 related columns in colData slot
col_data <- colData(sce_sc3)
head(col_data[ , grep("sc3_", colnames(col_data))])
```

```{r}
sce_sc3 <- sc3_calc_consens(sce_sc3)
```

```{r}
names(metadata(sce_sc3)$sc3$consensus)
```
## Find differentially expressed genes

```{r}
sce_sc3 <- sc3_calc_biology(sce_sc3)
```


## Inspect clusters

### Tables

```{r}
table(col_data$sc3_6_clusters)
```

```{r}
table(col_data$sc3_18_clusters)
```

```{r}
table(col_data$sc3_29_clusters)
```

### Visualize with Seurat

```{r}
# check that all cells are in the same order in both the sce_sc3 and Seurat objects
all(colnames(tnbc_seurat) == colnames(sce_sc3))
```

```{r}
tnbc_seurat$sc3_6_clusters <- sce_sc3$sc3_6_clusters
tnbc_seurat$sc3_18_clusters <- sce_sc3$sc3_18_clusters
tnbc_seurat$sc3_29_clusters <- sce_sc3$sc3_29_clusters
```

```{r}
DimPlot(tnbc_seurat, reduction = "umap",
		group.by = "seurat_clusters",
		label = TRUE, label.size = 5, repel = TRUE) +
	ggtitle("UMAP plot of Louvain clusters")
```

```{r}
DimPlot(tnbc_seurat, reduction = "umap",
		group.by = "lit_cell_type",
		label = TRUE, label.size = 5, repel = TRUE) +
	ggtitle("UMAP plot of Louvain clusters")
```

```{r}
DimPlot(tnbc_seurat, reduction = "umap",
		group.by = "sc3_6_clusters",
		label = TRUE, label.size = 5, repel = TRUE) +
	ggtitle("UMAP plot of SC3 clusters with k=6")
```

```{r}
DimPlot(tnbc_seurat, reduction = "umap",
		group.by = "sc3_18_clusters",
		label = TRUE, label.size = 5, repel = TRUE) +
	ggtitle("UMAP plot of SC3 clusters with k=18")
```
```{r}
DimPlot(tnbc_seurat, reduction = "umap",
		group.by = "sc3_29_clusters",
		label = TRUE, label.size = 5, repel = TRUE) +
	ggtitle("UMAP plot of SC3 clusters with k=29")
```
## Save updated SCE and Seurat objects

```{r}
save(tnbc_seurat, file = "tnbc_seurat_sc3.rda")
save(sce_sc3, file = "sce_sc3.rda")
```


